---
title: "Seeking Refuge in America: Understanding the Cap of Refugees Allowed in the US"
author: "Created by: Steven Herrera"
date: "5/01/2019"
output: html_document
---

### Loading packages

```{r load-packages, message=FALSE}
library(tidyverse)
library(infer)
library(openintro)
```

# Looking at the data

Originally, our data looked like this (showing only the first 10 results):

```{r}
refugee <- read.csv("refugee.csv")
refugee %>%
  head(10)
```

We had 39 observations, representing the years from 1980 to 2018. The variables in this dataset are `Year`, `Ceiling`, and `Admitted`.

1. **Year** (*numerical*) - represents the years from 1980-2018

2. **Ceiling** (*factor*) - represents the maximum amount of refugees accepted in those years.

3. **Admitted** (*factor*) - represents the actual amount of refugees accepted in those years.

In order to look at this data and make better use of the observations, I will turn the `Ceiling` and `Admitted` variables into numerical variables and remove the commas. Then, I will manipulate the data using the *mutate* and *filter* functions to double up on the observations and create a new variable called `Type` that, instead of having ceiling and admitted as variables, recognizes them as types with the numeric value (that would have gone under each respective variable) under the new variable `Toget`, or "together". Thus, our new dataset looks like this (showing only the first 10 results):

```{r}
refugee <- refugee %>%
  mutate(Ceiling = as.numeric(gsub(",", "", Ceiling)),
         Admitted = as.numeric(gsub(",", "", Admitted)))

only1 <- refugee %>% 
  mutate(Toget = Ceiling,
         Type = "Ceiling") %>%
  select(Year, Toget, Type)

only2 <- refugee %>%
  mutate(Toget = Admitted,
         Type = "Admitted") %>%
  select(Year, Toget, Type)

together <- rbind(only1, only2)

first_adm <- together %>%
  filter(Type == "Admitted") %>%
  head(5)
first_ceil <- together %>%
  filter(Type == "Ceiling") %>%
  head(5)

showing_first <- rbind(first_ceil, first_adm)
showing_first
```

Our new dataset `together` is now comprehensive and usable, in order to conduct analysis.

# Exploratory Data Analysis


Below, is a graph showing the changes in Ceiling and Admitted over the course of all years:

```{r}
ggplot(data = together, mapping = aes(x = Year, y = Toget, color = Type)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  labs(title = "Visualizing the differences between admitted and ceiling",
       subtitle = "Over the course of 1980-2018",
       x = "Year", y = "Number of refugees")
```

There have been significant changes since the 1980s and throughout the entire time period.

There have been some instances where the amount of refugees admitted went above the ceiling cap. Below, I showed which years those instances occurred:

```{r}
refugee %>%
  mutate(difference = Ceiling - Admitted) %>%
  filter(difference <= 0)
```

What happened in 1992 and 2017?

1. In 1992, according to the [Report of the United Nations High Commissioner for Refugees](https://www.unhcr.org/excom/unhcrannual/3ae68c860/report-united-nations-high-commissioner-refugees-1992.html), events that occurred in the Persian Gulf, the Horn of Africa and South-West Asia have posed unprecedented difficulties for people living in those areas, causing skyrocketing numbers of refugee applications for resettlement in the United States.

2. In 2017, according to this overview of [U.S. Refugee Law and Policy](https://www.americanimmigrationcouncil.org/research/overview-us-refugee-law-and-policy), current President Donald Trump signed an executive order that suspended the entire U.S. refugee admissions program for 120 days, which caused significant decreases in refugee applications. Regardless, the cap set on refugees wasn't strictly followed in 2018 since those admitted didn't go above the cap that year.

In being able to understand the influence the US has when setting caps on how many people are allowed to enter the US as refugees, it would be interesting and important to see if there a statistically significant difference between the amount of refugees coming to the US and the cap that's set on them. The reason why is because, although refugees have to send in an application from an overseas asylum officer and get it approved, exceptions can be made, often going over the cap set, which we saw it happen in 1992 and 2017. We'll conduct a hypothesis test to look at this.

# Hypothesis Test: Is there a difference?

### Do these data provide convincing evidence of a difference in refugees admitted for those that are type Ceiling and type Admitted during the years 1980s-2010s?

Let $mean$ represent the mean number of people admitted. Then,

$H_0: mean_{Ceiling} = mean_{Admitted}$  
$H_A: mean_{Ceiling} \ne mean_{Admitted}$

Below, I calculated the observed sample difference: 

```{r diff}
diff <- together %>%
  group_by(Type) %>%
  summarise(mean = mean(Toget))  %>%
  summarise(diff(mean)) %>%
  pull()
```

This value ended up being `r diff`. Then, we create a null distribution using 1000 permutations, in which we calculate the mean difference and record that on this distribution.

```{r}
null_dist <- together %>%
  specify(response = Toget, explanatory = Type) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "diff in means", order = c("Ceiling", "Admitted"))
```

Here, we have it visualized:

```{r}
ggplot(data = null_dist, aes(x = stat)) +
  geom_histogram() +
  geom_vline(xintercept = 14594.49, color = "red") +
  geom_vline(xintercept = -1 * 14594.49, color = "red") +
  labs(title = "Null distribution of differences in means",
       subtitle = "sampmed_Ceiling - sampmed_Admitted")
```

To determine the p-value, we calculate the proportion of permutation samples that yield a difference in sample means of -`r diff` or lower or `r diff` or higher.

```{r}
null_dist %>%
  filter(stat >= diff | stat <= -diff)  %>%
  summarise(pvalue = (n() / 1000))
```

This means that there isn't a statistically significant difference between those admitted and the cap that is set for all years. This is a good thing since we'd expect that the cap set does a good job at predicting how many people come to the U.S., but the implications behind such analysis also suggest that this might be obvious since the U.S. makes decisions based on policies and executive orders.

